Analysis-ready VCF at Biobank scale using Zarr

### 2025-01-30 response-to-reviewers.tex

\begin{point}
"There is currently no efficient, FAIR representation...". Just a nit and
feel free to ignore, but the solution you present is technically "current".
\end{point}
\begin{reply}
\end{reply}

\begin{point}
In Figure 2, the zarr line is occluded by the sav line and hard to see.
\end{point}
\begin{reply}
\end{reply}

\reviewersection
\section*{Reviewer 2 comments}

\begin{point}
The paper presents an encoding of the VCF data using Zarr to enable fast
retrieving subsets of the data. A vcf2arr conversion was provided and validated
on both simulated and real-world data sets. The topic of this work is
interesting and of good values, however, the experimental studies and
contributions should be considerable improved.
\end{point}
\begin{reply}
\end{reply}

\begin{point}
The proposed method is simply a conversion from VCF to Zarr format. Since
both are existing formats, the contributions and originality of this work are
not impressive.
\end{point}
\begin{reply}
\end{reply}

\begin{point}
The compression and query performance is the main concern of this work. The
method should be compared with other state-of-the-art queriable VCF compressors
like GTC, GBC, and GSC.
Danek A, Deorowicz S. GTC: how to maintain huge genotype collections in a
compressed form. Bioinformatics, 2018;34(11):1834-1840.
Zhang L, Yuan Y, Peng W, Tang B, Li MJ, Gui H,etal. GBC: a parallel toolkit
based on highly addressable byte-encoding blocks for extremely large-scale
genotypes of species. Genome Biology, 2023;24(1):1-22.
Luo X, Chen Y, Liu L, Ding L, Li Y, Li S, Zhang Y, Zhu Z. GSC: efficient
lossless compression of VCF files with fast query. Gigascience, 2024;
2;13:giae046. 
\end{point}
\begin{reply}
\end{reply}

\begin{point}
The method should be evaluated on more real VCF data sets.
\end{point}
\begin{reply}
\end{reply}

### 2025-01-31 paper.tex

Returning to Fig~\ref{fig-s3-throughput}, we can see that the af-dist
calculation is performed at a maximum rate of 25.5 GiB/s using 128
processes. At this rate, the af-dist calculation for the target 
% 180.80 MB in 4503 files
% Total data: 90060.00 MB in 4503 chnnks
region---172 MiB retrieved in 4503 chunks, 83.87 GiB of genotype data
decompressed and processed---
% NOTE: this doesn't quite up to the 25.5G above:  83.87 / 4.15 = 20.2
was completed in 4.15s.
This level of throughput would complete af-dist for all chromosomes for all samples
% compute: cost of c5.9xlarge per hour * 1.27
% ==                             $4.08 * 1.27
% io: ((total # variants / variant chunk size) * (sample count/sample chunk size)) /1000 * cost for 1000 IOPS
% ==  (((722000000       / 10000)              * (79000       /1000))              /1000)* $0.0004
in 1.27 hours for a compute cost of \$5.18 and an IO cost of \$2.28.
See \citep{durbin2020task,moore2021ome,gowan2022using} for further Zarr
benchmarks in cloud settings.

@@

\subsection{AWS parallel Zarr prototype}
This section describes the methodology used for the AWS parallel Zarr
throughput measurements in Fig~\ref{fig-s3-throughput}
and Fig~\ref{fig-s3-network-throughput}, and timing measurements
detailed in the cloud computing results.
These measurements were run on EC2 instances using Jupyter notebooks
hosted on Genomics England's Lifebit CloudOS platform. VCF Zarr chunks
were hosted in a standard profile S3 bucket. Both instance and bucket were
in the same availability zone in the eu-west-2 region. Each
of the parallel Python processes used the aioboto3 library (version 13.2.0) to
fetch chunks concurrently from S3.
% The underlying aiohttp library
% limits the concurrent fetching to 100 requests per process. 
For decompressing
and computing af-dist the same architecture was used but as fetched chunks
became available in memory in each process, they were decompressed and
processed synchronously. Decompression was performed by using the 
numcodecs library~\citep{numcodecs}.

>>>>

Returning to Fig~\ref{fig-s3-throughput}, we can see that the af-dist
calculation is performed at a maximum rate of 25.5 GiB/s using 128
processes. This level of throughput would complete af-dist for all
chromosomes for all samples (104 TiB of genotype data)
% timing: 722000000 * 79000 * 2bytes = 104 TiB
% 104TiB / 25.5GiB/s = 1.16 hours
% compute: cost of c5.9xlarge per hour * 1.16
% ==                             $4.08 * 1.16
% io: ((total # variants / variant chunk size) * (sample count/sample chunk size)) /1000 * cost for 1000 IOPS
% ==  (((722000000       / 10000)              * (79000       /1000))              /1000)* $0.0004
in 1.16 hours for a compute cost of \$4.73 and an IO cost of \$2.28.

Using this parallel architecture, the af-dist calculation for the target 
% 180.80 MB in 4503 files
% Total data: 90060.00 MB in 4503 chunks
region used for the single threaded benchmarks above ---172 MiB retrieved in 4503
chunks, 83.87 GiB of genotype data decompressed and processed---
% This gives an effective rate lower than  25.5G above:  83.87 / 4.15 = 20.2
was completed in 4.15s. A slightly slower effective throughput than in
Fig~\ref{fig-s3-throughput} as the short time to completion means that
the timing is sensitive to set-up and tear-down costs.
See \citep{durbin2020task,moore2021ome,gowan2022using} for further Zarr
benchmarks in cloud settings.

@@

\subsection{AWS parallel Zarr prototype}
This section describes the methodology used for the AWS parallel Zarr
throughput measurements in Fig~\ref{fig-s3-throughput}
and Fig~\ref{fig-s3-network-throughput}, and timing measurements
detailed in the cloud computing results.
These measurements were run on EC2 instances using Jupyter notebooks
hosted on Genomics England's Lifebit CloudOS platform. VCF Zarr chunks
were hosted in a standard profile S3 bucket. Both instance and bucket were
in the same availability zone in the eu-west-2 region. Each
of the parallel Python processes used the aioboto3 library (version 13.2.0) to
fetch chunks concurrently from S3. To measure throughput accurately, the
number of chunks fetched was scaled with the number of processes such that
each process fetched 60 GiB of decompressed data.
% The underlying aiohttp library
% limits the concurrent fetching to 100 requests per process. 
For decompressing
and computing af-dist the same architecture was used, but as fetched chunks
became available in memory in each process, they were decompressed and
processed synchronously. Decompression was performed by using the 
numcodecs library~\citep{numcodecs}.
